Containers and R---the Rockerverse and beyond

### 2019-11-06 manuscript.Rmd
restructure common envs section

One use for containers is to run on shared local hardware where teams manage their own high-performance servers.
This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
The former models provide modularity, while the latter approach is most similar to a simple shared server.
Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
The Rocker project provides containers pre-installed with the RStudio IDE.

In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.

In some cases, containers are not completely portable between hardware environments.
This is the case for software using graphical processing units (GPUs) which are increasingly popular for compute-intensive machine learning tasks.
Containers running GPU software  require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugn for docker and set of base images with appropriate drivers and libraries, and the Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages.

R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
This flexibility is even broader now that we have tools like `DBI`, that allows to create a common API for interfacing these databases, or like `dbplyr`, which are designed to run `dplyr` code straight against the database. 
But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by the IT. 
Giving access to a sandbox for the most common database environments is the idea behind [`r-db`](https://github.com/ColinFay/r-db) [@ColinFay], a docker image that contains everything needed to connect, from R, to a database. 
Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a docker instance of a specific database, and how to use `r-db` to connect to it. 
Each package contained inside this image also has a series examples, so that the user can get started with the database right away.

[RCloud](https://github.com/att/rcloud/tree/master/docker) uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.

>>>>

In **high-performance computing**, one use for containers is to run workflows on shared local hardware where teams manage their own high-performance servers.
This can follow one of several design patterns: users may deploy containers to hardware as a work environment for a specific project, conatiners may provide per-user persistent environments, or a single container can act as a common multi-user environment for a server.
In all cases, though, the containerized approach provides several advantages: First, users may use the same image and thus work environment on desktop and laptop computers, as well.
The former models provide modularity, while the latter approach is most similar to a simple shared server.
Second, software updates can be achieved by updating and redeploying the container, rather than tracking local installs on each server.
Third, the containerized environment can be quickly deployed to other hardware, cloud or local, if more resources are neccessary or in case of server destruction or failure.
In any of these cases, users need a method to interact with the containers, be it and IDE, or command-like access and tools such as SSH, which is usually not part of standard container recipes and must be added.
The Rocker project provides containers pre-installed with the RStudio IDE.
In cases where users store nontrivial amounts of data for their projects, data needs to persist beyond the life of the container.
This may be via in shared disks, attached network volumes, or in separate storage where it is uploaded between sessions.
In the case of shared disks or network-attached volumes, care must be taken to persist user permissions, and of course backups are still neccessary.
When working with multiple servers, an automation framework such as [Ansible](https://www.ansible.com) may be useful for managing users, permisions, and disks along with containers.

Using **GPUs** (graphical processing units) as a specialised hardware from containerized common work environments is also possible [REF].
GPUs are increasingly popular for compute-intensive machine learning tasks [REF].
Though in this case, containers are not completely portable between hardware environments.
Containers running GPU software require drivers and libraries specific to GPU models and versions, and containers require a specialized runtime to connect to the underlying GPU hardware.
For NVIDIA GPUs, the [NVIDIA Container Toolkit](https://github.com/NVIDIA/nvidia-docker) includes a specialized runtime plugin for Docker and a set of base images with appropriate drivers and libraries.
The Rocker project has (beta) images based on these that include GPU-enabled versions of machine-learning R packages: https://github.com/rocker-org/ml.

**Teaching** is a further example where sharing a prepared computing environment can greatly improve the process, especially for courses that require access to a relatively complex setup of software tools, e.g., as in the case of database systems.
R is a fantastic tool when it comes to interfacing with databases: almost every open source and proprietary database system has an R package that allows users to connect and interact with the it 
This flexibility is even broader now that we have tools like `DBI`, that allows to create a common API for interfacing these databases, or like `dbplyr`, which are designed to run `dplyr` code straight against the database. 
But learning and teaching these tools comes with a cost: the cost of deploying or having access to an environment with the softwares and drivers installed. 
For people teaching R, it can become a barrier if they need to install local versions of the drivers, or to connect to remote instances which might or might not be made available by IT services.
Giving access to a sandbox for the most common database environments is the idea behind **[`r-db`](https://github.com/ColinFay/r-db)** [@ColinFay], a Docker image that contains everything needed to connect, from R, to a database.
Notably, with `r-db`, the users don't have to install complex drivers or to configure their machine in a specific way. 
On top of that, `r-db` comes with a comprehensive [online guide](http://colinfay.me/r-db/), explaining how to spin a Docker instance of a specific database, and how to use `r-db` to connect to it.
Each package contained inside this image also has a series examples, so that the user can get started with the database right away.
The `rocker/tidyverse` base image ensures that users can also readily use packages for analysis, display, and reporting.

**[RCloud](https://github.com/att/rcloud/tree/master/docker)** uses a `rocker/drd` base image for creation of collaborative data analysis and visualisation environments.
